Assessing Factors Affecting Student's Performance during COVID-19 Pandemic

### 2024-02-15 outputs/paper/paper.qmd
Final Paper. All the changes are made

---
title: "My title"
subtitle: "My subtitle if needed"
author: 
  - First author
  - Another author
thanks: "Code and data are available at: LINK."
date: today
date-format: long
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
format: pdf
number-sections: true
bibliography: references.bib
---

@@

library(tidyverse) # a collection of data-related packages
library(knitr) # for kable
library(kableExtra)
library(here)
library(ggbeeswarm)
library(ggplot2)

#reading cleaned data

state_score = read_csv(file = here("outputs/data/cleaned_state_scoredata.csv"),       
  show_col_types = FALSE
)

head(state_score)

rhode = read_csv(
  file = here("outputs/data/cleaned_learningmodel/rhode.csv"),       
  show_col_types = FALSE
)

@@

@@

# Introduction

The COVID-19 pandemic has profoundly disrupted education systems worldwide, compelling schools and districts across the United States to rapidly adapt their learning models. In 2020 and 2021, educational institutions faced unprecedented challenges, transitioning between in-person, hybrid, and virtual learning models in response to evolving public health guidelines and infection rates. This shift has brought into sharp focus the need to understand the implications of these various schooling models on educational outcomes.

This paper delves into a comprehensive analysis across 11 U.S. states, examining the relationship between different schooling models and their impact on student enrollment and staffing patterns. The unique dataset compiled for this study encompasses a variety of metrics, including total enrollment, in-person and virtual attendance, as well as staff count across different learning models. By analyzing data from kindergarten to 12th grade, this study provides a granular view of how the pandemic has reshaped the educational landscape across these states.

Our findings reveal significant variations in how states have navigated the challenges posed by the pandemic, with notable differences in the adoption of in-person, hybrid, and virtual learning models. We observe that these variations are not only reflective of public health policies but also indicate broader socio-economic and demographic influences. For instance, preliminary analysis suggests that shifts to virtual learning correlate with changes in student enrollment, raising questions about equity and access in education during these challenging times.

The remainder of the paper is structured as follows: Section 2 provides a detailed overview of the data and methodology employed in this study. Section 3 presents an in-depth analysis of the schooling models across the 11 states, offering insights into the trends and patterns observed in the data. Section 4 discusses the key findings, exploring the implications of these schooling models on educational outcomes. Finally, Section 5 concludes with a reflection on the study's findings, limitations, and potential directions for future research in this critical area of educational policy.

The remainder of this paper is structured as follows. @sec-data....

# Data {#sec-data}

## Data Source and Collection

The data utilized in this study is obtained from the compilation of district-level schooling models and state standardized assessment data, originally featured in the paper "Pandemic Schooling Mode and Student Test Scores: Evidence from US School Districts" [@citepaper] published in the American Economic Review: Insights in June 2023. [@citewebsite]. The primary dataset encompasses schooling mode data from the 2020–2021 academic year across 11 U.S. states, integrating various educational approaches during the pandemic, including in-person, hybrid, and virtual learning environments. These datasets were sourced from COVID-19 School Data Hub [@citedata], district-level state standardized assessment data from spring
2016–2019 and 2021, and additional data demographic statistics from the National Center for Education Statistics (NCES) to establish a comprehensive analytical framework. We also used additional year 2020 household data by Household Pulse Survey [@citehousehold] from  U.S. Census Bureau spanning weeks 1 to 12.

**State Score Data**: This dataset encompasses state-level academic performance statistics from 2016 to 2020. Variables include the proportion of in-person, virtual, and hybrid learning, participation rates, standardized test pass rates, COVID-19 case rates per 100,000 in school zip codes, peak monthly cases, total enrollment figures, and political voting shares by district.

**Learning Model Data**: This dataset details the learning models adopted by school districts in Colorado, Connecticut, Ohio, Virginia, West Virginia, Wyoming, Mississippi, Rhode Island, Minnesota, Massachusetts, and Wisconsin. Each state's data outlines the specific learning model (in-person, virtual, hybrid) employed over time, enrollment figures, and staffing counts.

## Data Cleaning and Processing
We used R [@citeR] for data cleaning and processing, utilizing packages like tidyverse [@tidyverse] for data manipulation and janitor [@citejan] for cleaning column names. The cleaning process involved standardizing learning model categories, selecting columns of interest, and simplifying data for analysis. We integrated data from different states, and made sure that variations in reporting and measurement were addressed while combining them into one. For instance, some states categorized their schooling mode as "In-person," while others used "In-Person." To ensure they are equivalent, we standardized these terms. We also conducted a check for missing data values, and imputed or excluded them as appropriate based on the context. The variables were also renamed for clarity and consistency across different states' datasets.  A sample of cleaned state score data can be seen in [@tbl_cleaned_data].

```{r tbl_cleaned_data}
#| message: false
#| echo: false
#| warning: false
#| tbl-cap: Sample of Cleaned State Score Data
#| label: tbl_cleaned_data

library(kableExtra)
library(knitr)

```{r}
#| label: fig-bills
#| fig-cap: Bills of penguins
#| echo: false
#| warning: false

library(dplyr)

# Assuming each state data frame has the same structure and columns
all_states <- bind_rows(
  colorado %>% mutate(state = "Colprado"),
  connecticut %>% mutate(state = "Connecticut"),
  massachusets %>% mutate(state = "Massachusets"),
  minnesota %>% mutate(state = "Minnesota"),
  mississippi %>% mutate(state = "Mississsippi"),
  ohio %>% mutate(state = "Ohio"),
  rhode %>% mutate(state = "Rhode Islands"),
  virginia %>% mutate(state = "Virginia"),
  westvirginia %>% mutate(state = "West Virginia"),
  winscosin %>% mutate(state = "Winscosin"),
  wyoming %>% mutate(state = "Wyoming"),

)

state_score <- na.omit(state_score)

#therefore we scale it first, by taking proportions:
state_totals <- all_states %>%
  group_by(state) %>%
  summarise(total = n())

# Calculate the proportion of each learning model within each state
all_states <- all_states %>%
  group_by(state, learning_model) %>%
  summarise(count = n(), .groups = 'drop') %>%
  left_join(state_totals, by = "state") %>%
  mutate(proportion = count / total)

# Plotting with a bar graph
# ggplot(all_states, aes(x = state, y = proportion, fill = learning_model)) + 
#   geom_bar(stat = "identity", position = "dodge") +
#   theme_minimal() +
#   labs(x = "State", y = "Proportion", fill = "Learning Model", 
#        title = "Proportional Comparison of Learning Models in US States") +
  # scale_y_continuous(labels = scales::percent)

ggplot(all_states, aes(x = state, y = proportion, fill = learning_model)) + 
  geom_bar(stat = "identity", position = position_dodge(width = 0.9)) +
  theme_minimal() +
  labs(x = "State", y = "Proportion of Leaning Models", fill = "Learning Model", 
       title = "Proportional Comparison of Learning Models in US States") +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 8),
        plot.title = element_text(hjust = 0.5))

ggplot(state_score, aes(x = case_rate_per100k_zip, y = pass)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(x = "COVID-19 Case Rate per 100k", y = "Passing Rate", 
       title = "Correlation between Passing Rate and COVID-19 Case Rate per 100k") +
  theme_minimal() +
  xlim(0,100)

filtered_data <- state_score %>%
  filter(year %in% c(2019, 2021))

# Now create the scatter plot with the filtered data
ggplot(filtered_data, aes(x = case_rate_per100k_zip, fill = as.factor(year))) +
  geom_density(alpha = 0.5) +  # Adjust transparency with alpha
  labs(x = "COVID-19 Case Rate per 100k", y = "Density",
       title = "Density Plot of COVID-19 Case Rates per 100k (2019 vs 2021)",
       fill = "Year") +
  theme_minimal()

# Create a box plot
state_score <- state_score %>% 
  filter(!is.na(case_rate_per100k_zip) & !is.nan(case_rate_per100k_zip))

# Create a box plot with colors
ggplot(state_score, aes(x = category, y = pass, fill = category)) +
  geom_boxplot() +
  labs(title = "Box Plot of Passing Rates by COVID-19 Case Rate Categories", 
       x = "COVID-19 Case Rate Category", 
       y = "Passing Rate") +
  theme_minimal() +
  scale_fill_brewer(palette = "Pastel1")

ggplot(state_score, aes(x = pass)) +
  geom_density(fill = "blue", alpha = 0.5) +
  labs(title = "Density Plot of Passing Rates", x = "Passing Rate", y = "Density") +
  theme_minimal()

```
## Data Modifications

In this study, we created a new dataset by carefully selecting and combining data from the Household Pulse Survey, available from the U.S. Census Bureau for weeks 1 to 12. This process involved choosing key variables that were crucial for our research. We aimed to analyze the connection between economic changes, particularly the loss of employment income, and their impact on educational outcomes, focusing specifically on school passing rates. This helped us to understand the relationship between these economic factors and educational performance. To ensure consistency, we standardized terminologies and measurements, and aligned variables such as timeframes and demographic categories across the datasets. For example, we combined the state score data with the weekly household data, aligned variables relevant to our study. The merged data file was thereafter saved as a CSV file to study the trends later on.

# Results
In the Results section of our study, our analysis discusses two key geographical representations:

1. Passing Rate Trends Over Different Years and Subjects: Figure [@fig] shows the decline in passing rates in English Language Arts (ELA) and Mathematics from 2016 to 2021, depicting a consistent downward trend across both subjects especially during pandemic period from 2019 to 2021.

2. Variations in Passing Rates Between Different States: The box plot comparison [@fig2] across states showcases the wide range of passing rates, with some states showing a wide spread, which indicates the  variability in performance within those states.

Building on these overviews, we visualized the distribution of learning models across states using box plots. The box plot for in-person learning [@box_plot_inperson] revealed a range of engagement across states, with Virginia showing a median of 0 percentage in-person learning during 2019 to 2021, whereas Wyoming having a median of approximately 80 percent in-person learning.

The virtual learning model box plot [@box_plot_virtual] showed a significant number of outliers, suggesting that certain states had districts with a notably high share of virtual learning.The states Wyoming and Rhode Island had a median of 0 percent virtual learning whereas Virginia had highest median of virtual learning.  Lastly, the hybrid learning model [@box_plot_hybrid] box plot displayed variability, with some states having a wide interquartile range, showing a diverse adoption of this learning model. However, Colorado had a median of 0 percent adoption of hybrid learning model relative to other states. These visual insights set the foundation for further statistical examination in subsequent analyses.

After assessing the distribution of in-person, virtual, and hybrid learning models across states, we narrowed our focus to further investigate the impact of virtual learning, as this model showed a significant variation and prevalence during the pandemic period. We plotted a scatter plot [@scatterplot] and examined the relationship between the share of virtual learning and student passing rates. The scatter plot indicates a negative trend, which implies that an increase in the share of virtual learning correlates with a decrease in passing rates. The data points suggest that higher virtual learning is associated with lower academic performance of students.

After the initial analysis of learning model, we further studied the impact of Covid-19 on student's learning. We again plotted box plots [@covidboxplot] to observe the relationship between COVID-19 case rates per 100,000 and passing rates. COVID-19 case rates per 100,000 were categorized into 'Low,' 'Medium,' and 'High' based on the data's quartile distribution. It is revealed from the plot that districts with 'Low' COVID-19 case rates showed a median passing rates near the 50th percentile, while those categorized as 'High' showed a decline in median passing rates.

We further expanded our analysis to examine the relationship between the average number of students per staff and the average pass rate for virtual learners across the 11 states [@fig]. Due to insufficient data, Rhode Island (RI) and Mississippi (MS) were not graphed. The graph plots both these variables and shows an overall positive correlation between the number of students per staff on average vs the average pass rate. The trend indicates that the states with the higher pass rates also had more students assigned to a teacher. We note that Virginia (VA) had the highest student-to-staff ratio, while Wyoming (WY) had the lowest. While this may be not expected of a general learning environment, the virtual learning model may be a factor contributing to this positive relationship between the student-to-staff ratio and the average pass rates across the states.

```{r}
#| label: box_plot_inperson
#| fig-cap: Relationship between wing length and width
#| echo: false
#| warning: false
#| message: false

library(dplyr)

@@

# Step 2: Plot the data
ggplot(inperson_data, aes(x = state, y = share_inperson)) +
  geom_boxplot() +
  theme_minimal() +
  labs(title = "Box Plot of In-Person Learning Model Share Across States",
       x = "State", y = "Share of In-Person Learning") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

virtual_data <- state_score %>%
  select(state, share_virtual)

# Plot for virtual learning model
ggplot(virtual_data, aes(x = state, y = share_virtual)) +
  geom_boxplot() +
  theme_minimal() +
  labs(title = "Box Plot of Virtual Learning Model Share Across States",
       x = "State", y = "Share of Virtual Learning") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Prepare data for hybrid learning model
hybrid_data <- state_score %>%
  select(state, share_hybrid)

# Plot for hybrid learning model
ggplot(hybrid_data, aes(x = state, y = share_hybrid)) +
  geom_boxplot() +
  theme_minimal() +
  labs(title = "Box Plot of Hybrid Learning Model Share Across States",
       x = "State", y = "Share of Hybrid Learning") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggplot(state_score, aes(x = share_virtual, y = pass)) +
  geom_point() +  # Add points to the plot
  geom_smooth(method = "lm", se = FALSE) +  # Add a linear trend line without the shaded confidence interval
  theme_minimal() +
  labs(title = "Scatter Plot of Virtual Learning Share vs. Passing Rate",
       x = "Share of Virtual Learning", y = "Passing Rate") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

state_with_highest_virtual <- state_score %>%
  group_by(state) %>%
  summarize(average_virtual = mean(share_virtual, na.rm = TRUE)) %>%
  arrange(desc(average_virtual)) %>%
  slice(1) %>%
  pull(state)

# Filter the data for the state with the highest virtual learning
state_passing_data <- state_score %>%
  filter(state == state_with_highest_virtual)

# Plot the passing rates against the case rate per 100k for the state with the highest virtual learning
ggplot(state_passing_data, aes(x = case_rate_per100k_zip, y = pass)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +  # Add a linear trend line
  theme_minimal() +
  labs(title = paste("Passing Rate vs. Case Rate per 100k in", state_with_highest_virtual),
       x = "Case Rate per 100k", y = "Passing Rate") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

state_with_highest_cases <- state_score %>%
  group_by(state) %>%
  summarize(average_cases = mean(case_rate_per100k_zip, na.rm = TRUE)) %>%
  arrange(desc(average_cases)) %>%
  slice(1) %>%
  pull(state)

# Filter the data for the state with the highest case rate per 100k
state_passing_data <- state_score %>%
  filter(state == state_with_highest_cases)

# Plot the passing rates for the state with the highest case rate per 100k
ggplot(state_passing_data, aes(x = as.factor(year), y = pass, group = state, color = state)) +
  geom_line() +
  geom_point() +
  theme_minimal() +
  labs(title = paste("Passing Rate in", state_with_highest_cases, "with the Highest COVID-19 Case Rate"),
       x = "Year", y = "Passing Rate") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Find the state with the lowest case rate per 100k
state_with_lowest_cases <- state_score %>%
  group_by(state) %>%
  summarize(average_cases = mean(case_rate_per100k_zip, na.rm = TRUE)) %>%
  arrange(average_cases) %>%
  slice(1) %>%
  pull(state)

# Filter the data for the state with the lowest case rate per 100k
state_passing_data_lowest <- state_score %>%
  filter(state == state_with_lowest_cases)

# Plot the passing rates for the state with the lowest case rate per 100k
ggplot(state_passing_data_lowest, aes(x = as.factor(year), y = pass, group = state, color = state)) +
  geom_line() +
  geom_point() +
  theme_minimal() +
  labs(title = paste("Passing Rate in", state_with_lowest_cases, "with the Lowest COVID-19 Case Rate"),
       x = "Year", y = "Passing Rate") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

@@

# Model

The goal of our modelling strategy is twofold. Firstly,...

Here we briefly describe the Bayesian analysis model used to investigate... Background details and diagnostics are included in [Appendix -@sec-model-details].

## Model set-up

Define $y_i$ as the number of seconds that the plane remained aloft. Then $\beta_i$ is the wing width and $\gamma_i$ is the wing length, both measured in millimeters.

\begin{align} 
y_i|\mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \\
\mu_i &= \alpha + \beta_i + \gamma_i\\
\alpha &\sim \mbox{Normal}(0, 2.5) \\
\beta &\sim \mbox{Normal}(0, 2.5) \\
\gamma &\sim \mbox{Normal}(0, 2.5) \\
\sigma &\sim \mbox{Exponential}(1)
\end{align}

We run the model in R [@citeR] using the `rstanarm` package of @rstanarm. We use the default priors from `rstanarm`.

### Model justification

We expect a positive relationship between the size of the wings and time spent aloft. In particular...

We can use maths by including latex between dollar signs,  instance $\theta$.

Our results are summarized in @tbl-modelresults.

>>>>

---
title: "Assessing Factors Affecting Student's Performance during COVID-19 Pandemic"
subtitle: "Data from USA"
author: 
  - Vanshika Vanshika 
  - Shivank Goyal
  - Navya Hooda
thanks: "Code and data are available at: https://github.com/shivankgoel003/Student-Performance-COVID-19-US-States-Factors; Minimal and Full Replication on Social Science Reproduction platform available at [https://www.socialsciencereproduction.org/reproductions/1470/published/index](https://www.socialsciencereproduction.org/reproductions/1470/published/index) "
date: today
date-format: long
abstract: "We analyze data from 11 U.S. states to assess the correlation between factors that may be indicative of student performance during COVID-19 learning disruptions. This paper replicates data analysis upon pandemic schooling and its affect on student test scores, and  further attempts to understand other mechanisms that led to learning losses during this period. Several charesteritics were related to student performance: mode of learning, loss of COVID-19 case rates, employment and student-to-staff ratio.The learning losses were found to be around 13–14 percentage points in math and approximately 8 percentage points in ELA between 2019 and 2021 (Jack and Oster 2023). In particular we discover that  all factors except student-to-staff ratio were relevant predictors and contributed to noticeable differences when compared to 2019 (pre-covid), these findings suggest that the factors that normally contribute to student performance had exacerbated effects during COVID-19 as they led to a larger magnitude of learning losses."
format: pdf
number-sections: true
bibliography: references.bib
toc: true
---

@@

library(tidyverse) # a collection of data-related packages
library(knitr) # for kable
library(kableExtra)
library(here)
library(ggbeeswarm)
library(reshape2)
library(ggplot2)
library(dplyr)

#reading cleaned data
rhode = read_csv(
  file = here("outputs/data/cleaned_learningmodel/rhode.csv"),       
  show_col_types = FALSE
)

@@

scores = read_csv(
  file = here("outputs/data/all_scoredata.csv"),       
  show_col_types = FALSE
)
scores_2019 = read_csv(
  file = here("outputs/data/score_ohio.csv"),       
  show_col_types = FALSE
)

jobs = read_csv(
  file = here("outputs/data/combined_cleaned_data.csv"),       
  show_col_types = FALSE
)

dataset1  = read_csv(
  file = here("outputs/data/cleaned_state_scoredata.csv"),       
  show_col_types = FALSE
)
state_score  = read_csv(
  file = here("outputs/data/cleaned_state_scoredata.csv"),       
  show_col_types = FALSE
)
merged_data = read_csv(file = here("outputs/data/merged_data_by_state.csv"), show_col_types = FALSE)
```
# Introduction

The COVID-19 pandemic has been one of the most challenging times globally in recent decades. The measures taken in place to reduce the transmission across the world were unprecedented but necessary in curbing the spread of disease. However, the cost of these measures posed disruption in every part of life whether it be businesses, healthcare, or education. For young people in particular, COVID-19 posed disruptions and threatened their sense of normalcy. The transition to COVID-19 learning manifested an environment of uncertainty and stress at a time of critical psychological development [@impact].

The COVID-19 pandemic has created the largest disruption of education systems in history, affecting nearly 1.6 billion learners in more than 190 countries and all continents [@un]. COVID-19 led to rapid adaptations in learning models across schools and districts in the United States. From 2020 to 2021, educational institutions faced unprecedented challenges, transitioning between in-person, hybrid, and virtual learning models in response to evolving public health guidelines and infection rates. This shift in learning models has had severe impacts on student performance unlike anything we’ve seen in recent years. As reported by NAEP, from 2019 to 2022, the average performance on the National Assessment of Educational Progress [@naep] declined by 8 points in 8-th grade math, representing approximately 75% of a grade level, and by 3 points in eighth-grade reading, equivalent to roughly 25% of a grade level. While the pandemic was bound to impact learning outcomes of students, its likely that various factors played a role in the overall results, and not just the pandemic alone.

In normal times, student performances have been studied and analyzed to highlight strong indicators of student performance. Among these important factors affecting students’ academic achievement are their economic background and conditions, personal health conditions, and teacher involvement. In particular, it is suggested that the adequacy of the physical conditions that make up the school, such as the information technology facilities, location, library and media facilities, and schoolyard are attributes important role in the student’s academic achievement. (@teachers) In the context of COVID-19 schooling, all these factors were impacted in some way. The physicality of a school environment was shifted to a virtual hybrid space, students’ health whether their own or of those around them was impacted due to the transmission of COVID-19, the pandemic brought in employment losses, and teachers were expected to rapidly provide learning in new formats. This changing dynamic raises the question of how these factors impacted student performance during COVID-19.

In this paper, we embark on a comprehensive analysis to explore the estimand whether factors like loss of employment, COVID case rates, student-to-staff ratio in virtual learning, and the mode of learning had a significant impact on student performance, measured by the pass rates of students in grades 3-8  across the 11 U.S states namely Ohio (OH), Colorado (CO), Connecticut (CT), Minnesota (MN), Massachusetts (MA), Mississippi(MS), Rhode Island (RI), Virginia (VA), Wyoming (WY), Wisconsin (WI), and West Virginia (WV). 
 Through examination, we show the transformative impact of the pandemic on the educational landscape across these states. Our analytical framework explores whether the patterns of the factors stated above contributed to the decline in test scores between 2019 and 2021, shedding light on the nuanced associations between the two.

We explore the origin of this decline in detail by using the extensive dataset from the Household Pulse Survey [@citehousehold], searching for household dynamics that may have caused students’ lower academic performance and states’ diminished passing rates. Our findings reveal that the learning losses found during the pandemic were both substantial and highly heterogeneous across communities. We claim that these losses cannot be attributed to one cause but rather stem from a variety of diverse factors as we stated previously. Notably, the loss of household income during the pandemic exhibited a direct correlation with learning loss, with mathematics suffering a more significant decline compared to ELA courses. Additionally, we also explored the correlation between passing rates in each state and the type of learning mode used popularly,  which allowed us to demonstrate the varying impacts of in-person, hybrid, or virtual models on student pass rates. Furthermore, we highlight the significant effect of varying COVID-19 case rates on learning outcomes, in general, we notice that states experiencing higher infection rates manifested greater learning losses compared to their counterparts with lower case counts.

Recognizing the intricate effects of the pandemic on education is vital for crafting effective strategies moving forward. By grasping the various factors affecting student learning amidst these challenges, we can provide schools and districts with the tools they need to overcome obstacles and ensure every student receives the necessary support to succeed academically and beyond. We have the opportunity to strengthen the education system to counteract learning setbacks and pave the way for a brighter future for all students.

This paper is structured as follows. In the Data Section, we denote the sources of the datasets utilized in the paper we are replicating, the specific techniques we used to process them, and highlight their key factors. In the Results Section we dive into the findings we discovered following replication and workings of the original paper, we assessed how they were relevant to our study of the factors contributing to the decline in student performance. In the Discussion Section, we address biases and weaknesses in the data that contribute to our findings, and how we approached analyzing those areas despite the limitations.

# Data {#sec-data}
## Data Source and Collection

The data utilized in this study is obtained from the compilation of district-level schooling models and state standardized assessment data, originally featured in the paper "Pandemic Schooling Mode and Student Test Scores: Evidence from US School Districts" [@citepaper] published in the American Economic Review: Insights in June 2023. [@citewebsite]. The primary dataset encompasses schooling mode data from the 2020–2021 academic year across 11 U.S. states, integrating various educational approaches during the pandemic, including in-person, hybrid, and virtual learning environments. These datasets were sourced from COVID-19 School Data Hub [@citedata], district-level state standardized assessment data from spring
2016–2019 and 2021, and additional data demographic statistics from the National Center for Education Statistics (@nces) to establish a comprehensive analytical framework. We also used additional year 2020 household data by Household Pulse Survey [@citehousehold] from  U.S. Census Bureau spanning weeks 1 to 12.

**State Score Data**: This dataset encompasses state-level academic performance statistics from 2016 to 2020. Variables include the proportion of in-person, virtual, and hybrid learning, participation rates, standardized test pass rates, COVID-19 case rates per 100,000 in school zip codes, peak monthly cases, total enrollment figures, and political voting shares by district. These variables are measured by the standardized test scores of each state. Each state records their scores differently and on different websites, so the methodology of measurements is unknown.

**Learning Model Data**: This dataset details the learning models adopted by school districts in Colorado, Connecticut, Ohio, Virginia, West Virginia, Wyoming, Mississippi, Rhode Island, Minnesota, Massachusetts, and Wisconsin. Each state's data outlines the specific learning model (in-person, virtual, hybrid) employed over time, enrollment figures, and staffing counts. These variables are measured by NCES Demographic Data. Both district- and school-level demographic data from the National Center for Education Statistics Common Core of Data (https://nces.ed.gov/ccd/)
were accessed through Urban Institute’s Education Data Portal using Stata 16 (https://educationdata.urban.org/documentation/). Data documentation can be found at the website above. COVID-19 Community Case Rate Data: We included COVID-19 Community Case Rate Data by District from the COVID-19 School Data Hub website accessed through the “Download Community Case
Rate Data” in the Community Case Rate Data by District box on the 'For Researchers' page of the CSDH website (https://www.covidschooldatahub.com/for_researchers).

## Data Cleaning and Processing
We used R [@citeR] for data cleaning and processing, utilizing packages like tidyverse [@tidy] for data manipulation and janitor [@jan] for cleaning column names. Other packages used includes `ggplot2` [@ggplot], `dplyr` [@dp], `readr` [@read], `tibble` [@tib], `janitor` [@jan],`reshape2` [@reshape], `knitr` [@knit], `ggbeeswarm` [@gg], `ggrepel` [@repel], `kableExtra`[@kable] and `here` [@here].The cleaning process involved standardizing learning model categories, selecting columns of interest, and simplifying data for analysis. We integrated data from different states, and made sure that variations in reporting and measurement were addressed while combining them into one. For instance, some states categorized their schooling mode as "In-person," while others used "In-Person." To ensure they are equivalent, we standardized these terms. We also conducted a check for missing data values, and imputed or excluded them as appropriate based on the context. The variables were also renamed for clarity and consistency across different states' datasets.  A sample of cleaned state score data can be seen in [@tbl-cleaned-data].
```{r}
#| label: tbl-cleaned-data
#| message: false
#| echo: false
#| warning: false
#| tbl-cap: "Sample of Cleaned State Score Data"

### Student-to-Staff Ratio Across the 11 States

This variable was not directly provided in our original paper, to retrieve this relationship we 
required 2 variables, average pass rates across each state, and the average student-to-staff ratio for each state. To find these we had to use 2 different types of datasets to finally combine them to reach our variable of interest.
We first used “cleaned_state_scoredata.dta” in .csv format, this dataset contains pass rates for various districts across the 11 U.S. states. Since we require an average across the states we grouped all the state pass rates in 2021 for each state by filtering and calculated it by dividing the pass rate sum by the number of total pass rates reported in one specific state, this yields the average pass rate for each state as needed. We then proceed to find the student-to-staff ratio. This information is specific to each state and found in the replication data. We use each state’s xlsx file, which contains the variables ‘enrollment_total’ and ‘staff_count’. The statistics are reported biweekly in this data, so we choose to aggregate them by finding all unique values for enrollment and teacher numbers, with that we find the ratio by dividing the total number of students by staff count. We do this calculation for each state by looping through each file and yielding a simple average for each state are newly created and cleaned results seen in [@tbl-teacher].

```{r}
#| label: tbl-teacher
#| tbl-cap: "Sample of Cleaned Student-to-Staff Ratio"
#| message: false
#| echo: false
#| warning: false

options(kableExtra.latex.max.width=1)

# Define the data
state <- c("CO", "CT", "MA", "MN", "MS", "OH", "RI", "VA", "WI", "WV", "WY")
average_pass <- c(32.3143154318402, 50.2760636293435, 43.862676056338, 46.0322019263922, 59.9536587026782, 58.8878278763826, 34.7934722039435, 54.1523990342111, 38.6599490324205, 32.0722727272727, 51.9001212765773)
average_students_per_teacher <- c(6.54852303777137, 9.75566450808731, 6.73622340678238, 7.32408458156817, NA, 4.85888729378285, NA, 11.1500196439537, 6.55786715773129, 6.19012111835609, 3.57947403685482)

# Create the data frame
df <- data.frame(State = state, `Pass Rate` = average_pass, `Average Students per Teacher` = average_students_per_teacher)

# Print the table
df %>%
  kable(col.names = c("State", "Pass Rate", "Avg. Students per Teacher"), 
        format = "latex",
        booktabs = TRUE) %>%
  kable_styling(latex_options = c("striped", "scale_down"), font_size = 6)
```

## Data Modifications

In this study, we developed a unique dataset by carefully selecting and combining data from the Household Pulse Survey [@citehousehold], available from the U.S. Census Bureau for weeks 1 to 12. This data was measured by using the survey to get both qualitative and quantitative data. This process involved choosing key variables that were crucial for our research. We aimed to analyze the connection between economic changes, particularly the loss of employment income, and their impact on educational outcomes, focusing specifically on school passing rates. This helped us to understand the relationship between these economic factors and educational performance. To ensure consistency, we standardized terminologies and measurements, and aligned variables such as timeframes and demographic categories across the datasets. For example, we combined the state score data with the weekly household data, aligned variables relevant to our study. The merged data file was thereafter saved as a CSV file to study the trends later on.

```{r}
#| label: fig-marks
#| fig-cap: "Comparison between Maths and ELA "
#| echo: false

# Comparing passing rates between different subjects
subject_comparison <- ggplot(scores, aes(x = subject, y = pass, fill = subject)) +
  geom_boxplot() +
  labs(title = "Passing Rates Comparison Between Different Subjects",
       x = "Subject",
       y = "Passing Rate") +
  theme_minimal()

# Display the graph
subject_comparison

@@

Here are two comprehensive overviews detailing passing rate trends. Furthermore, we delve into the factors contributing to these trends.

1. Passing Rate Trends Over Different Years and Subjects: [@fig-subject] shows the decline in passing rates in English Language Arts (ELA) and Mathematics from 2016 to 2021, with larger decline in maths than ELA [@fig-marks],  depicting a consistent downward trend across both subjects especially during the pandemic period from 2019 to 2021.
```{r}
#| label: fig-subject
#| fig-cap: "Passing Rate Trend"
#| echo: false

# new graph
passing_rate_trends <- ggplot(scores, aes(x = year, y = pass, color = subject)) +
  geom_line() +
  labs(title = "Passing Rate Trends Over Different Years and Subjects",
       x = "Year",
       y = "Passing Rate",
       color = "Subject") +
  theme_minimal()
passing_rate_trends
```
2. Variations in Passing Rates Between Different States: The box plot comparison [@fig-passing-trend] across states showcases the wide range of passing rates, with some states showing a widespread, which indicates the variability in performance within those states.

```{r}
#| label: fig-passing-trend
#| fig-cap: "Variations in Passing Rates Between Different States"
#| echo: false

# Variations 
passing_rate_variations <- ggplot(scores, aes(x = state, y = pass, fill = state)) +
  geom_boxplot() +
  labs(title = "Variations in Passing Rates Between Different States",
       x = "State",
       y = "Passing Rate") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
passing_rate_variations

## Impact of Learning Model
As the COVID-19 pandemic changed the educational model, schools across the United States adopted various learning methods to continue educating students such as virtual, hybrid, and in-person learning. This segment of our study focuses on how these learning models were implemented across different states and their impact on the academic performance of students.

We visualized the distribution of learning models across states using box plots. The box plot for in-person learning ,[@fig-plot-inperson], reveals a wide range of engagement across states, with Virginia showing a median of 0 percent in-person learning from 2019 to 2021, whereas Wyoming had a median of approximately 80 percent in-person learning.

```{r}
#| label: extra
#| fig-cap: "set up for graphs"
#| echo: false
#| warning: false
#| message: false
#| results: hide

library(patchwork)

@@

virtual_data <- state_score %>%
  select(state, share_virtual)
# Prepare data for hybrid learning model
hybrid_data <- state_score %>%
  select(state, share_hybrid)

@@

```{r}
#| label: fig-plot-inperson
#| fig-cap: "In-Person Learning Model"
#| echo: false
#| warning: false
#| message: false
# Step 2: Plot the data
ggplot(inperson_data, aes(x = state, y = share_inperson)) +
  geom_boxplot() +
  theme_minimal() +
  labs(title = "In-Person Learning Model Share Across States",
       x = "State", y = "Share of In-Person Learning") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```{r}
#| label: fig-plot-virtual
#| fig-cap: "Virtual Learning Model"
#| echo: false
#| warning: false
#| message: false
# Plot for virtual learning model
ggplot(virtual_data, aes(x = state, y = share_virtual)) +
  geom_boxplot() +
  theme_minimal() +
  labs(title = "Virtual Learning Model Share Across States",
       x = "State", y = "Share of Virtual Learning") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Prepare data for hybrid learning model
hybrid_data <- state_score %>%
  select(state, share_hybrid)
```

```{r}
#| label: fig-plot-hybrid
#| fig-cap: "Hybrid Learning Model"
#| echo: false
#| warning: false
#| message: false
# Plot for hybrid learning model
ggplot(hybrid_data, aes(x = state, y = share_hybrid)) +
  geom_boxplot() +
  theme_minimal() +
  labs(title = "Hybrid Learning Model Share Across States",
       x = "State", y = "Share of Hybrid Learning") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

The virtual learning model box plot [@fig-plot-virtual] showed a significant number of outliers, suggesting that certain states had districts with a notably high share of virtual learning. The states Wyoming and Rhode Island had a median of 0 percent virtual learning whereas Virginia had the highest median of virtual learning.
Lastly, the hybrid learning model [@fig-plot-hybrid] box plot displayed variability, with some states having a wide interquartile range, indicating a diverse adoption of this learning model. However, Colorado had a median of 0 percent adoption of the hybrid learning model relative to other states. These visual insights set the foundation for further statistical examination in subsequent analyses.

After assessing the distribution of in-person, virtual, and hybrid learning models across states, we narrowed our focus to further investigate the impact of virtual learning, as this model showed a significant variation and prevalence during the pandemic period. We plotted a scatter plot [@fig-scatterplot] and examined the relationship between the share of virtual learning and student passing rates. The scatter plot indicates a negative trend, which implies that an increase in the share of virtual learning correlates with a decrease in passing rates. The data points suggest that higher virtual learning is associated with lower academic performance of students.
